---
title: "Too hard to get: the role of probabilistic expectations and cognitive complexity in destructive multi-dimensional reference points"
shorttitle: "Destructive multi-dimensional reference points"

author:
  - name: Aspen Han
    corresponding: true
    orcid: 0000-0003-1474-7968
    email: xiangyuhan@uchicago.edu
    affiliations:
      - name: University of Chicago
        department: Department of Economics, Division of Social Sciences
        address:  1126 E. 59th Street
        city: Chicago
        region: IL
        postal-code: 60637

author-note:
  disclosures:
    study-registration: null
    data-sharing: null
    related-report: null
    conflict-of-interest: null
    financial-support: null
    authorship-agreements: null
    gratitude: null
    
abstract: "This is the abstract. Lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum lorem ipsum"

keywords: [reference points, multi-dimensional, conflicting]

bibliography: bibliography.bib

floatsintext: true

editor_options: 
  chunk_output_type: console

knitr:
  opts_chunk:
    out.width: "75%"
    
format:
  apaquarto-pdf:
    documentmode: man
    keep-tex: true
    list-of-figures: true
    fig-width: 6
    fig-asp: 0.618
    fig-align: center
    include-in-header:
      text: |
      #remove extra space between main list and sublist
        \usepackage{enumitem}
        \setlist[itemize]{topsep=0pt,itemsep=0pt,partopsep=0pt,parsep=0pt}
        \setlist[enumerate]{topsep=0pt,itemsep=0pt,partopsep=0pt,parsep=0pt}
  
---

```{r}
#| label: setup
#| include: false

# Load required packages; install them with command below if not already done so, where package name is the argument to the library function
# install.package("<package name>")

## Data reading and writing, transformation, and visualisation
library(tidyverse)
library(readxl) #read excel data
library(reshape2) #alternative to tidyr though it is retired
library(broom) #turns function output into tidy tibbles


## Statistical inference and regression analysis
library(car) #basic regression analysis
library(sandwich) #estimate robust errors
library(estimatr) #run robust and other more complex regressions and analysis
library(nnet) #run multinomial logistic regressions
library(fixest) #run fixed effects regressions
library(lmtest) #run diagnostic tests on linear models

## Tables
library(texreg) #make regression tables
library(gtsummary) #make summary stats tables
library(gt) #make general tables

#Alternatively, can use load packages using require(), e.g.
# require(tidyverse)

#Note that using require() will throw a warning but still execute code afterwards, which can create further errors, whereas using library() will throw a warning and stop running the code, which is preferred

```

# 1. Introduction



# 2. Methodology

I outline my identification strategy to answer my research question in three parts

* Theoretical specification
* Experimental design
* Context, sample, and data

## A. Theoretical specification



## B. Experimental design



## C. Context, sample, and data



# 3. Analysis and results

I provide an overview of key features in the experimental data before delving deeper into probability analysis to answer my research questions.

## A. Data overview

```{r}
#| label: Default-plot-themes

#Set default theme for ggplot; serif is times new roman

def_theme <- theme(plot.title = element_text(family = "serif", size = 15, hjust = 0.5),
                  axis.title = element_text(family = "serif", size = 13),
                  axis.text = element_text(family = "serif", size = 11),
                  strip.text.x = element_text(family = "serif", size = 12),
                  strip.text.y = element_text(family = "serif", size = 12, angle = 0),
                  legend.title = element_text(family = "serif", size = 12),
                  legend.text = element_text(family = "serif", size = 11))
```

## B. Probability analysis



# 4. Discussion



## Relationship with covariates



## Robustness checks



## Limitations



# 5. Conclusion



\newpage

# Bibliography

::: {#refs}
:::

\newpage

# Appendix

## I. Derivation of reference-dependent on-the-job search model